Nengo: A Python tool for building large-scale functional brain models

### 2013-10-03 nengo.tex
Added references

Modeling the human brain
is one of the greatest
scientific challenges of our time.
Computational neuroscience
has made significant advancements;
however, we still lack a mathematical account of
how biological and electrical components
implement cognitive functions such as sensory processing,
memory formation, reasoning, and motor control.
Much work has been put into neural simulators
that attempt to recreate neuroscientific
data in precise detail with the thought that
cognition will emerge by connecting
detailed neuron models according
to the statistics of biological synapses \cite{TODO}. % blue brain
Cognition has not yet emerged
from data-driven large scale models.
On the other end of the spectrum,
cognitive architectures \cite{TODO} % act-r
and theoretical frameworks \cite{TODO} % Hawkins, Kurzweil
have solved high-dimensional statistical problems,
but do so without respecting biological constraints,
and with worse performance than machine learning approaches.

Nengo is a neural simulator
based on a theoretical framework proposed
by Eliasmith \& Anderson \citeyearpar{TODO}
called the Neural Engineering Framework
(NEF).
The NEF is a large-scale modeling approach
that can leverage any neuron model
to build neural networks with
demonstrable cognitive abilities.
Nengo and the NEF has been used to build
increasingly sophisticated neural subsystems
for the last decade (e.g., \cite{TODO}),
culminating recently with Spaun,
currently the world's
largest functional brain model \cite{TODO}.
Spaun is a network of 2.5 million spiking neurons
that can perform eight high-level cognitive tasks
including memorizing lists and inductive reasoning.
It can perform any of these eight tasks
at any time by being presented
the appropriate series of images
representing the task to be performed
and the details of that task;
for example, sequentially presenting images
containing the characters $A3[1234]$ instructs Spaun
to memorize the list $1234$.
If asked to recall the memorized list,
it sends motor commands to a simulated arm model,
tracing out the digits $1234$.
While the tasks Spaun performs are diverse,
all tasks use a common set of
functional modules.
Each functional module corresponds
to a brain area that has been hypothesized
to perform those functions
in neuroscientific literature.

@@

JavaNengo is a suite of tools composed of
a neural simulator and graphical user interface (GUI)
implemented in Java,
and a scripting interface implemented in Jython
(as detailed in Table~\ref{tab:nengo-vers}).
JavaNengo's simulator was created
with the intention of being a general
neural simulator that included
the methods of the NEF
as one of several
methods for creating neural models.
As a result, its implementation
of important NEF components
are nested in deep class hierarchies,
adding unnecessary layers of complexity
for developers wishing
to extend those NEF components.
For example,
the most often used class for making
NEF models, \texttt{NEFEnsemble},
extends from five ancestor classes
and implements seven interfaces.
Adding support for synaptic plasticity
in \texttt{NEFEnsemble}
involved adding or significantly modifying
25 Java classes \cite{TODO}.

@@

The JavaNengo scripting interface,
described in \cite{TODO},
provides a simpler interface
to the NEF-specific portions of the simulator.
This greatly enhanced productivity,
allowing models to be described
by short scripts.
Modellers quickly started using
the scripting interface
as their primary tool,
but still needed to access the GUI
for interactive model inspection,
and still needed to modify the simulator
to add new functionality
(e.g., to add new neuron models).
This is because Jython makes it easy
to inspect and create Java objects,
but does not make it easy to
write Java code to
inspect and create Jython objects.
Modellers, therefore, had to be
proficient in Java whether
they used the scripting interface or not.
As there are few neural modellers
that are proficient with both Java and Python syntax,
the scripting interface still lacks
convenient ways to record the results of simulations,
and to define complex experimental setups.
Additionally, Jython is always several versions
behind CPython, making it impossible
to use new language features
like dictionary comprehensions.
Jython also cannot access libraries
that contain any C code,
including all scientific libraries
that use NumPy.

@@

PyNengo is a Python package for
defining and simulating
neural models using the NEF.
Its interfaces are based on NumPy,
so it integrates seamlessly
with other scientific Python tools
for interactive exploration \cite{TODO}, %Ipython
scientific computation \cite{TODO}, %scipy
and plotting \cite{TODO}. % Matlab
Arbitrary CPython programs
can use Nengo models,
opening up possibilities
for using neurally implemented algorithms
in web services, games, and other applications.

PyNengo has a simple object model,
which makes it easy to
document, test, and modify.
These objects are accessible
through a scripting interface
similar to JavaNengo's \cite{TODO}, % frontiers 2009
with improvements
for easier incorporation of user-defined networks,
neuron models and learning rules.
Critically, and unlike JavaNengo,
PyNengo's scripting interface
contains functions to record simulated data
and access that data after simulation.

@@

The Neural Engineering Framework (NEF; \cite{TODO})
proposes three principles
that enable the construction
of large-scale neural models.

@@

\paragraph{Representation}
Information is encoded by populations of neurons.
We represent information
with time-varying vectors of real numbers
in order to manipulate that information
using conventional mathematics.
By doing this, we can \textit{encode}
those vectors by injecting
specific amounts of current into
single neuron models based on
the vector being encoded.
This drives the neuron,
causing it to spike.
With enough neurons,
we can estimate the originally encoded vector
through a \textit{decoding} process.
This idea is similar to population coding
\cite{TODO}, but generalized
to vectors of arbitrary length.

@@

An \texttt{Ensemble} is
a population of neurons
that represents information
in the form of a real-valued vector.
When creating an ensemble,
the user must provide the dimensionality
(i.e., the length of the vector it represents),
and an object that describes
a population of neurons.
For example, \texttt{Ensemble('A', nengo.LIF(50, tau\_ref=0.002), 1)}
describes an ensemble representing a one-dimensional vector
with 50 leaky integrate-and-fire neurons \cite{TODO} % LIF cite
that have a refractory period of 2 milliseconds.
The \texttt{nengo.LIF} class defines
the parameters of the LIF neurons symbolically
so that each simulator can compute
the LIF nonlinearity efficiently.
The neuron model used by the ensemble
is changed by passing in a different symbolic neuron object;
however, the simulator used must be aware
of that type of neuron.

@@

Decoupling model creation and simulation
has been done previously
by PyNN \cite{TODO}.
In PyNN, the same Python script
can be used to run a model
on four different simulators.
Nengo follows this programming model by
decoupling model creation and simulation,
which enables PyNengo simulators
to allocate memory and schedule computations
in the most efficient ways possible.
Simulators are given a \texttt{Model}
as an argument;
this \texttt{Model} is a symbolic description,
which a simulator can assume will not change.
The simulator, however,
can modify the model as it sees fit;
in most cases, this means that the simulator
will fill in many of the details
not specified at the symbolic level.

@@

\begin{figure}
\begin{center}
  \includegraphics[width=.85\textwidth]{comm_channel}
  \begin{minipage}{0.48\textwidth}
    \begin{lstlisting}
import nengo
from nengo.helpers import white_noise
model = nengo.Model("Communication Channel")
model.make_node("Input", white_noise(1, 5))
model.make_ensemble("A", nengo.LIF(30), 1)
model.make_ensemble("B", nengo.LIF(30), 1)
model.connect("Input", "A")
model.connect("A", "B")
model.probe("Input")
model.probe("A", filter=0.01)
model.probe("A.spikes")
model.probe("B", filter=0.01)
model.probe("B.spikes")
sim = model.simulator()
sim.run(1)
A_data = sim.data("A")
A_spikes = sim.data("A.spikes")
...
    \end{lstlisting}
  \end{minipage}
  \begin{minipage}{0.39\textwidth}
    \includegraphics[width=\textwidth]{comm_channel_res}
  \end{minipage}
\end{center}
 \textbf{\refstepcounter{figure}\label{fig:comm-channel}
   Figure \arabic{figure}.}{
   A communication channel implemented with Nengo.
   (A) Diagram depicting the model. Ensemble A
   projects its encoded signal to ensemble B unchanged.
   (B) Nengo code to build and simulate the model
   for 1 second.
   (C) The results of the simulation.
   The input signal (top panel) is white noise limited to 0 to 5 Hz.
   The signal is well represented by both ensemble A (middle panel)
   and ensemble B (bottom panel) despite the neural firing patterns
   (underlaid in middle and bottom panels) being different.}
\end{figure}

@@

Many models in theoretical neuroscience
are based on attractor networks \cite{TODO}. % Amit, Rolls, Deco
The NEF has been used in the past
to implement many different types of
attractor networks \cite{TODO}.
Figure~\ref{fig:lorenz} depicts
a Nengo implementation of the Lorenz chaotic attractor
with a single ensemble
composed of 2000 leaky integrate-and-fire neurons.
We have implemented the Lorenz attractor
in PyNN for benchmarking purposes
(code not shown).

@@

The communication channel and Lorenz attractor
are small models that demonstrate
the principles of the NEF.
Their small size enables us to write
PyNN scripts that implement roughly
the same functionality
with Brian, NEURON, NEST, and TODO PCSim.
We run each parameter set five times
on the same machine,
and plot the mean.
In most cases, the coefficient of variation
is well below 0.1, except for two
outliers with coefficients of 0.18 and 0.22,
indicating that these means are robust.
The results, shown in Figure~\ref{fig:benchmarks}A and B,
suggest that all versions of Nengo are significantly
faster than the simulators accessible
through PyNN, especially
as the size of models increase.
This is likely due to Nengo's
use of factorized weight matrices,
rather than storing and computing with
the entire weight matrix
on each timestep.
While NEST and NEURON were not
run with MPI,
the reference simulator of PyNengo
also only uses one CPU core.
The results also suggest that PyNengo's
simulators are faster than JavaNengo's simulator.

@@

The most closely related project is PyNN
\cite{TODO},
which provides a high-level scripting
interface similar to the
high-level object model in PyNengo.
Rather than implementing its
own simulator, PyNN uses four existing
simulators with Python bindings,
such as NEST \cite{TODO}.

>>>>

Modeling the human brain
is one of the greatest
scientific challenges of our time.
Computational neuroscience
has made significant advancements;
however, we still lack a mathematical account of
how biological and electrical components
implement cognitive functions such as sensory processing,
memory formation, reasoning, and motor control.
Much work has been put into neural simulators
that attempt to recreate neuroscientific
data in precise detail with the thought that
cognition will emerge by connecting
detailed neuron models according
to the statistics of biological synapses \citep{markram2006}.
Cognition has not yet emerged
from data-driven large scale models.
On the other end of the spectrum,
cognitive architectures \citep{anderson2004}
and theoretical frameworks \citep{hawkins2006}
have solved high-dimensional statistical problems,
but do so without respecting biological constraints,
and with worse performance than machine learning approaches.

Nengo is a neural simulator
based on a theoretical framework proposed
by \citet{eliasmith2003}
called the Neural Engineering Framework
(NEF).
The NEF is a large-scale modeling approach
that can leverage any neuron model
to build neural networks with
demonstrable cognitive abilities.
Nengo and the NEF has been used to build
increasingly sophisticated neural subsystems
for the last decade
(e.g., path integration [\citealp{conklin2005}],
working memory [\citealp{singh2006}],
list memory [\citealp{choo2010}],
inductive reasoning [\citealp{rasmussen2011}],
motor control [\citealp{dewolf2011}],
decision making [\citealp{stewart2012}])
culminating recently with Spaun,
currently the world's
largest functional brain model \citep{eliasmith2012}.
Spaun is a network of 2.5 million spiking neurons
that can perform eight high-level cognitive tasks
including memorizing lists and inductive reasoning.
It can perform any of these eight tasks
at any time by being presented
the appropriate series of images
representing the task to be performed
and the details of that task;
for example, sequentially presenting images
containing the characters $A3[1234]$ instructs Spaun
to memorize the list $1234$.
If asked to recall the memorized list,
it sends motor commands to a simulated arm model,
tracing out the digits $1234$.
While the tasks Spaun performs are diverse,
all tasks use a common set of
functional modules.
Each functional module corresponds
to a brain area that has been hypothesized
to perform those functions
in neuroscientific literature.

@@

JavaNengo is a suite of tools composed of
a neural simulator and graphical user interface (GUI)
implemented in Java,
and a scripting interface implemented in Jython
(as detailed in Table~\ref{tab:nengo-vers}).
JavaNengo's simulator was created
with the intention of being a general
neural simulator that included
the methods of the NEF
as one of several
methods for creating neural models.
As a result, its implementation
of important NEF components
are nested in deep class hierarchies,
adding unnecessary layers of complexity
for developers wishing
to extend those NEF components.
For example,
the most often used class for making
NEF models, \texttt{NEFEnsemble},
extends from five ancestor classes
and implements seven interfaces.
Adding support for synaptic plasticity
in \texttt{NEFEnsemble}
involved adding or significantly modifying
25 Java classes \citep{bekolay2010}.

@@

The JavaNengo scripting interface,
described in \citet{stewart2009},
provides a simpler interface
to the NEF-specific portions of the simulator.
This greatly enhanced productivity,
allowing models to be described
by short scripts.
Modellers quickly started using
the scripting interface
as their primary tool,
but still needed to access the GUI
for interactive model inspection,
and still needed to modify the simulator
to add new functionality
(e.g., to add new neuron models).
This is because Jython makes it easy
to inspect and create Java objects,
but does not make it easy to
write Java code to
inspect and create Jython objects.
Modellers, therefore, had to be
proficient in Java whether
they used the scripting interface or not.
As there are few neural modellers
that are proficient with both Java and Python syntax,
the scripting interface still lacks
convenient ways to record the results of simulations,
and to define complex experimental setups.
Additionally, Jython is always several versions
behind CPython, making it impossible
to use new language features
like dictionary comprehensions.
Jython also cannot access libraries
that contain any C code,
including all scientific libraries
that use NumPy.

@@

PyNengo is a Python package for
defining and simulating
neural models using the NEF.
Its interfaces are based on NumPy,
so it integrates seamlessly
with other scientific Python tools
for interactive exploration \citep{perez2007},
scientific computation \citep{oliphant2007},
and plotting \citep{hunter2007}.
Arbitrary CPython programs
can use Nengo models,
opening up possibilities
for using neurally implemented algorithms
in web services, games, and other applications.

PyNengo has a simple object model,
which makes it easy to
document, test, and modify.
These objects are accessible
through a scripting interface
similar to JavaNengo's \citep{stewart2009},
with improvements
for easier incorporation of user-defined networks,
neuron models and learning rules.
Critically, and unlike JavaNengo,
PyNengo's scripting interface
contains functions to record simulated data
and access that data after simulation.

@@

The Neural Engineering Framework (NEF; \citealp{eliasmith2003})
proposes three principles
that enable the construction
of large-scale neural models.

@@

\paragraph{Representation}
Information is encoded by populations of neurons.
We represent information
with time-varying vectors of real numbers
in order to manipulate that information
using conventional mathematics.
By doing this, we can \textit{encode}
those vectors by injecting
specific amounts of current into
single neuron models based on
the vector being encoded.
This drives the neuron,
causing it to spike.
With enough neurons,
we can estimate the originally encoded vector
through a \textit{decoding} process.
This idea is similar to population coding
\citep{georgopoulos1986}, but generalized
to vectors of arbitrary length.

@@

An \texttt{Ensemble} is
a population of neurons
that represents information
in the form of a real-valued vector.
When creating an ensemble,
the user must provide the dimensionality
(i.e., the length of the vector it represents),
and an object that describes
a population of neurons.
For example, \texttt{Ensemble('A', nengo.LIF(50, tau\_ref=0.002), 1)}
describes an ensemble representing a one-dimensional vector
with 50 leaky integrate-and-fire neurons \citep{lapicque1907}
that have a refractory period of 2 milliseconds.
The \texttt{nengo.LIF} class defines
the parameters of the LIF neurons symbolically
so that each simulator can compute
the LIF nonlinearity efficiently.
The neuron model used by the ensemble
is changed by passing in a different symbolic neuron object;
however, the simulator used must be aware
of that type of neuron.

@@

Decoupling model creation and simulation
has been done previously
by PyNN \citep{davison2008}.
In PyNN, the same Python script
can be used to run a model
on four different simulators.
Nengo follows this programming model by
decoupling model creation and simulation,
which enables PyNengo simulators
to allocate memory and schedule computations
in the most efficient ways possible.
Simulators are given a \texttt{Model}
as an argument;
this \texttt{Model} is a symbolic description,
which a simulator can assume will not change.
The simulator, however,
can modify the model as it sees fit;
in most cases, this means that the simulator
will fill in many of the details
not specified at the symbolic level.

@@

\begin{figure}
\begin{center}
  \includegraphics[width=.73\textwidth]{comm_channel}
  \begin{minipage}{0.37\textwidth}
    \begin{lstlisting}[basicstyle={\footnotesize\ttfamily}]
import nengo
from nengo.helpers import white_noise
model = nengo.Model("Communication Channel")
model.make_node("Input", white_noise(1, 5))
model.make_ensemble("A", nengo.LIF(30), 1)
model.make_ensemble("B", nengo.LIF(30), 1)
model.connect("Input", "A")
model.connect("A", "B")
model.probe("Input")
model.probe("A", filter=0.01)
model.probe("A.spikes")
model.probe("B", filter=0.01)
model.probe("B.spikes")
sim = model.simulator()
sim.run(1)
A_data = sim.data("A")
A_spikes = sim.data("A.spikes")
...
    \end{lstlisting}
  \end{minipage}
  \begin{minipage}{0.3\textwidth}
    \includegraphics[width=\textwidth]{comm_channel_res}
  \end{minipage}
\end{center}
 \textbf{\refstepcounter{figure}\label{fig:comm-channel}
   Figure \arabic{figure}.}{
   A communication channel implemented with Nengo.
   (A) Diagram depicting the model. Ensemble A
   projects its encoded signal to ensemble B unchanged.
   (B) Nengo code to build and simulate the model
   for 1 second.
   (C) The results of the simulation.
   The input signal (top panel) is white noise limited to 0 to 5 Hz.
   The signal is well represented by both ensemble A (middle panel)
   and ensemble B (bottom panel) despite the neural firing patterns
   (underlaid in middle and bottom panels) being different.}
\end{figure}

@@

Many models in theoretical neuroscience
are based on attractor networks \citep{amit1992, deco2003}.
The NEF has been used in the past
to implement many different types of
attractor networks \citep{eliasmith2005}.
Figure~\ref{fig:lorenz} depicts
a Nengo implementation of the Lorenz chaotic attractor
with a single ensemble
composed of 2000 leaky integrate-and-fire neurons.
We have implemented the Lorenz attractor
in PyNN for benchmarking purposes
(code not shown).

@@

The communication channel and Lorenz attractor
are small models that demonstrate
the principles of the NEF.
Their small size enables us to write
PyNN scripts that implement roughly
the same functionality
with Brian \citep{goodman2009}, NEURON \citep{hines2009},
NEST \citep{eppler2008}, and PCSim \citep{pecevski2009}.
We run each parameter set five times
on the same machine,
and plot the mean.
In most cases, the coefficient of variation
is well below 0.1, except for two
outliers with coefficients of 0.18 and 0.22,
indicating that these means are robust.
The results, shown in Figure~\ref{fig:benchmarks}A and B,
suggest that all versions of Nengo are significantly
faster than the simulators accessible
through PyNN, especially
as the size of models increase.
This is likely due to Nengo's
use of factorized weight matrices,
rather than storing and computing with
the entire weight matrix
on each timestep.
While NEST and NEURON were not
run with MPI,
the reference simulator of PyNengo
also only uses one CPU core.
The results also suggest that PyNengo's
simulators are faster than JavaNengo's simulator.

@@

The most closely related project is PyNN
\citep{davison2008},
which provides a high-level scripting
interface similar to the
high-level object model in PyNengo.
Rather than implementing its
own simulator, PyNN uses four existing
simulators with Python bindings,
such as NEST.
